Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-14 paper.tex
Added protein+tunnel images
added table
abstract, intro

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}}

@@

\title*{
    Testing traversability of protein tunnels with flexible ligands using motion planning
}

\author{
    Vojt\v ech Von\' asek \and 
    Barbora Kozl\'\i kov\'a  \and 
    Adam Jur\v{c}\'\i k  \and 
    David Bend\'{a}\v{r}  \and
    Martin Saska }

\institute{
    V. Von\'{a}sek, Martin Saska \at 
    Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}  
\and
B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno, \\
\and 
\inst{3}
TODO DAVID \at TODO DAVID
}

@@

\abstract{
\red{needs rewrite:
Function and behavior of proteins are achieved by interactions with other molecules such as ligands.
The interaction takes place in an active site inside the protein, which is accessible from the surface by so called tunnels.
The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
behavior of the protein.
Widely used tools for tunnel detection are based on Voronoi diagrams and they compute the tunnels for a virtual spherical probe.
The tunnels are then scored based on their bottleneck, length and also curvature. 
It is however not easy to determine if a given non-spherical ligand can pass such a tunnel.
To provide the chemists better information about the tunnels, the full shape of ligands should be considered.
}
Sampling-based motion planning techniques can be used to compute trajectories for the non-spherical ligands in the protein tunnels.
In this paper, we propose a modification of Rapidly Exploring Random Tree (RRT) to find the trajectories.
To enable the ligands to move in the tunnels, the planning is realized with a reduced-scale of the ligands. 
The flexibility of ligands is achieved by using a predefined set of ligand conformations.
For the purpose of further analysis, we computed trajectories are visualized ..
}

\section{Introduction \& motivation}

@@

\red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}

%Protein structures are essential components of all living organisms.
Proper understanding of structure and function of proteins is important in many fields, including drug design, agriculture, cosmetics, etc.
%Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
Computational methods are often used to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
from the protein surface by a tunnel.

The knowledge about the tunnels is important in many fields..
For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
ligand from the outside environment to the active site or vice versa.

@@

To present the results to chemicists ...

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
  \item \textbf{Ligand scale} --- The scale of the ligand wheh it passes a tunnel's part.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
\red{ i.e., the conditional probability we discussed in the mail?}
\end{itemize}
During visualization, we map these properties onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
\quad & \quad
\includegraphics[width=0.3\textwidth]{fig/throughput}
\end{tabular}
\caption{Tunnel properties mapped onto its surface.
Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
\label{fig:properties}
\end{figure}

\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }

Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
In order to visualize the important trajectories we first cluster the trajectories according to their similarity. 
\red{can you please add some details how is the similarity counted? Maybe a short description of even equation.}
Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
\red{please some reference to the hierarchical clustering}.
Moreover, we map the representant's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.45\textwidth]{fig/trajectories-all}
\qquad & \qquad
\includegraphics[width=0.45\textwidth]{fig/trajectories-clustered-21} \\
\end{tabular}
\caption{Visualization of resulting trajectories.
Left: All trajectories colored according to whether they reached the end of the tunnel (green) or they did not (red).
Right: Trajectories clustered according to their proximity colored by distinct colors.
The different line widths represent the number of trajectories in a cluster.}
\label{fig:trajectories}
\end{figure}

@@

The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
The tunnel is represented by 760 spheres with average distance $0.05$~\AA.

% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).

\begin{figure}
\centering
\caption{\label{fig::tunnel}
    TODO image of the tunnel + 1cqw
}
\end{figure}

\begin{table}
\centering
\caption{\label{tab::rrt}
    Runtime and sucess ratio of the planner
}
\input table.003.tex
\hskip 10pt
%\input table.004.tex
\end{table}

Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.

\section{Conclusion }

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}

@@

\title*{
    Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning
}

\author{
    Vojt\v ech Von\' asek \and 
    Barbora Kozl\'\i kov\'a  \and 
    Adam Jur\v{c}\'\i k  \and 
    David Bedn\'{a}\v{r}  \and
    Martin Saska }

\institute{
    V. Von\'{a}sek, Martin Saska \at 
    Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}  
\and
B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno, \\
\and 
\inst{3}
David Bedn\'{a}\v{r} \at 
Faculty of Science,
Masaryk University,
Kamenice 5, 625 00 Brno
}

@@

\abstract{
Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
Current methods describing the ligand passing through the tunnel are based either only on geometry of the tunnel or on molecular dynamic simulations. 
In this paper, we propose \red{TODO}
Additionally, we introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
On the other hand, it is much less computationally demanding than methods based on molecular dynamic simulations. 
The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
}

\section{Introduction \& Motivation}

@@

Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.

For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
ligand from the outside environment to the active site or vice versa.

@@

To present the results to chemists ...

@@

\section{Conclusion }
In the future, we will focus on the following extensions proposed by the domain experts.
Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
